---
title: Flipping your Way to Freedom
author: ~
date: '2020-05-05'
slug: flipping-your-way-to-freedom
categories: []
tags: [riddler, puzzles]
subtitle: 'Simulations, logic, and some calculus!'
summary: ""
authors: []
lastmod: '2020-05-05T08:46:45-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

In this post, we will solve a classic puzzle from the
[Riddler](https://fivethirtyeight.com/features/can-you-flip-your-way-to-freedom/). We will use R to simulate the puzzle and build some intuition, and then use good old calculus to compute a closed form solution to the problem!

## Problem statement

<blockquote>
From Bart Wright comes a rhetorical question from a famed soliloquy, “To flip, or not to flip?”:

You are locked in the dungeon of a faraway castle with three fellow prisoners
(i.e., there are four prisoners in total), each in a separate cell with no means
of communication. But it just so happens that all of you are logicians (of
course).

To entertain themselves, the guards have decided to give you all a single chance
for immediate release. Each prisoner will be given a fair coin, which can either
be fairly flipped one time or returned to the guards without being flipped. If
all flipped coins come up heads, you will all be set free! But if any of the
flipped coins comes up tails, or if no one chooses to flip a coin, you will all
be doomed to spend the rest of your lives in the castle’s dungeon.

The only tools you and your fellow prisoners have to aid you are random number
generators, which will give each prisoner a random number, uniformly and
independently chosen between zero and one.

What are your chances of being released?

**Extra credit:** Instead of four prisoners, suppose there are N prisoners. Now
what are your chances of being released?

</blockquote>


## Simulate the problem

<span class='marginnote'>
This blog uses http://hypothes.is for enabling inline comments. You can
highlight any line of text and click to annotate. You will need to sign up
for an account with hypothes.is. You can also use this to take private notes!
</span>
Let us first try to understand the state of policies available to the prisoners.
Since the only mechanism available is a random number generator, any policy
would have to depend on the number generated. Moreover, the prisoner would need
to decide whether or not to flip a coin based on this number.

Without any loss of generality, we can assume that prisoner $i$ decides to flip
a coin if the random number generated is less than $\theta_i$. Since the
prisoners cannot communicate with each other, the only optimal policy that
exists is $\theta_1 = \theta_2 = ... \theta$.


```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
source('~/Downloads/theme-rv.R')
theme_set(theme_rv_light(base_family = 'Alegreya Sans'))
rv_set_chunk_opts()
```

We can now simulate the impact of this policy on the probability of freedom as a
function of the number of prisoners, and the threshold $\theta$.

``` {r simulate-policy}
simulate_policy <- function(theta = 0.5, nb_prisoners = 4){
  # Generate random numbers for all prisoners
   x <- runif(nb_prisoners)
   # Toss if x > theta: N = No Toss, H = Head, T = Tail
   tosses <- ifelse(x < theta, "N", sample(c("H", "T"), nb_prisoners, replace = TRUE))
   # Freedom: #H > 0 and #T == 0 
   sum(tosses == "H") > 0 && sum(tosses == "T") == 0
}
mean(replicate(1000, simulate_policy()))
```

Now that we have a function to simulate the policy, we can use it to compute the outcome for a number of trials across different values of `theta` and `nb_prisoners`. The easiest way to do this is to create a grid of values using `expand.grid` and apply the function `simulate_policy` to compute the outcome of the trial.

```{r simulations, eval = FALSE}
NB_TRIALS = 10^4
simulations <- expand.grid(
  id_trial = 1:NB_TRIALS,
  theta = seq(0, 1, 0.01),
  nb_prisoners = 4:10
) %>%
  mutate(is_free = map2_lgl(theta, nb_prisoners, simulate_policy))

simulations
```

```{r simulations-head, echo = FALSE}
simulations <- readr::read_rds('simulations.rds')
head(simulations)
```

We can summarize the simulated outcomes across `theta` and `nb_prisoners` and 
plot them.

```{r outcomes-plot, warning = FALSE, message = FALSE, fig.show = 'hide'}
outcomes <- simulations %>%
  group_by(theta, nb_prisoners) %>%
  summarize(prob_free = mean(is_free))

outcomes %>%
   mutate(nb_prisoners = factor(nb_prisoners)) %>%
   ggplot(aes(x = theta, y = prob_free, color = nb_prisoners)) +
   geom_smooth(se = FALSE, span = 0.2) +
   geom_point(alpha = 0.5) +
   geom_hline(yintercept = 0.25, linetype = 'dashed', color = 'gray20') +
   guides(colour = guide_legend(nrow = 1)) +
   labs(
     title = "Probability of freedom",
     subtitle = "",
     x = "Probability of NOT flipping a coin",
     y = "",
     color = "Number of prisoners"
   ) +
   theme(plot.title.position = 'plot')
```

<span class='marginnote' style='margin-top:1em;'>
Note that I use `geom_smooth` to plot the curve and overlay the points on top of
it. This provides a better approximation of behavior since we are only
simulating N = 10^4 trials.
</span>

<span class='marginnote' style='margin-top:1em;'>
😀 You can click on any of the plots to view an enlarged version!
</span>

```{r outcomes-plot-print, ref.label = "outcomes-plot", echo = FALSE}

```


We can now recover the approximate probabilities of
freedom and plot it as a function of the number of prisoners. Note how the
probabilities seem to be really close to each other and around 25%.


```{r probability-across-n}
outcomes %>%
  group_by(nb_prisoners) %>%
  filter(prob_free == max(prob_free)) %>%
  ungroup() %>%
  ggplot(aes(x = nb_prisoners, y = prob_free)) +
  geom_col(fill = 'purple', color = NA) +
  geom_hline(yintercept = 0.25, linetype = 'dashed') +
  labs(
    title = "Probability of freedom",
    subtitle = "",
    x = 'Number of prisoners',
    y = ''
  ) +
  theme(plot.title.position = 'plot')
```
## Exact solution

Maybe, we can solve this problem exactly without simulations. Let **N** denote
the event that a prisoner does not flip the coin, and $\theta$ be the
probability of this event. The only scenario under which the prisoners end up
being free is when __none__ of them toss a tail, and __all__ of them don't end 
up NOT tossing. Hence, we can now compute the probability of freedom as:


$$ 
\begin{aligned}
P(Freedom) &= P(T^cT^cT^c...T^c) - P(NNN...N) \\
&= \Big(\frac{1+\theta}{2}\Big)^n - \theta^n 
\end{aligned}
$$

Now that we have a closed form expression for the probability of freedom, we can
turn this into an R function and use the `optimize()` function to compute the 
optimal values.


```{r optimize-probability}
prob_freedom <- function(n = 4){
  function(theta){
   ((1 + theta)/2)^n - theta^n
  }
}
optimize(prob_freedom(4), c(0, 1),  maximum = TRUE)
```

We can extend this approach to compute the optimal probability of freedom
across a range of values for the number of prisoners.

<span class='marginnote'>
We use `unnest_longer` to unpack the list of values returned by `optimize()`.
The `tidyr` package is full of such utility functions!
</span>


```{r optimal-values}
optimal_values <- data.frame(nb_prisoners = 1:30) %>%
  mutate(prob_free = map(nb_prisoners, ~ {
    optimize(prob_freedom(.x), c(0, 1),  maximum = TRUE)
  })) %>%
  tidyr::unnest_longer(prob_free)
head(optimal_values)
```

```{r optimized-plot, fig.show = 'hide'}
optimal_values %>% 
  ggplot(aes(x = nb_prisoners, y = prob_free, color = prob_free_id)) +
  geom_line(size = 1.5) +
  geom_hline(yintercept = c(0.25, 0.96), linetype = 'dashed', color = 'gray40') +
  scale_color_discrete(
    labels = c("Not Flipping Coin", "Freedom")
  ) +
  labs(
    title = "Probability of freedom",
    subtitle = "",
    x = 'Number of prisoners',
    y = 'Probability',
    color = ""
  ) +
  theme(
    plot.title.position = 'plot'
  )
```

<span class='marginnote' style='margin-top:1rem;'>
It seems like the probabilities of freedom, and NOT flipping the coin
converge asymptotically to the values 0.25 and 0.96 respectively. It would
be interesting if we can prove this result mathematically!
</span>

```{r echo = FALSE, ref.label = 'optimized-plot'}

```

Let us take the derivative of the probability function and see if we can derive
a closed form expression for the optimal value of $\theta$, by setting the 
derivative to zero. Let $P(\theta)$ represent the probability of freedom. Recall
that we can express it as:

$$
P(\theta) = \Big(\frac{1 + \theta}{2}\Big)^{n} - \theta^n
$$
The first derivative can be written down as:

$$
P^{'}(\theta) = 2^{n} n(1 + \theta)^{n - 1} - n\theta^{n-1}
$$

Solving for $P^{'}(\theta) = 0$, we get

$$
\theta_{opt} = \Big[\exp\big(\frac{n\log(2)}{n - 1}\big) - 1\Big]^{-1}
$$

Voila! That is a really beautiful result!! Plugging the optimal value of
$\theta$, we can write down the probability of freedom as:

$$
P(\theta_{opt}) = \Big[\exp\big(\frac{n\log(2)}{n - 1}\big) - 1\Big]^{-n}
$$

Let us now use this exact expression and plot the probabilities of freedom and 
not deciding to toss the coin.


```{r closed-form, echo = FALSE}
f <- function(n){
  theta = (exp(n*log(2) / (n - 1)) - 1)^-1
  prob_free = prob_freedom(n)(theta)
  list(maximum = theta, objective = prob_free)
}

tibble(n = 2:20) %>%
  mutate(f = map(n, f)) %>%
  unnest_longer(f) %>%
  ggplot(aes(x = n, y = f, color = f_id)) +
  geom_line(size = 1.5) +
  geom_hline(yintercept = c(0.25, 0.96), linetype = 'dashed', color = 'gray40') +
  scale_color_discrete(
    labels = c("Not Flipping Coin", "Freedom")
  ) +
  ylim(0, 1) +
  labs(
    title = "Probablity of freedom",
    subtitle = 'These values are calculated using the closed form expression derived.',
    x = 'Number of prisoners',
    y = '',
    color = ""
  )
```


<script type="application/json" class="js-hypothesis-config">
{
  "theme": "clean"
}
</script>
<script async src="https://hypothes.is/embed.js"></script>
